Bayesian Deep Gaussian Processes for Correlated Functional Data: \\ A Case Study in Cosmological Power Spectra

### 2025-05-06 paper/main.tex
annie done in section 2

Notice, in the center panel of Figure \ref{fig:plot_data}, the precision of each
spectra increases as $k$ increases.  The curves get progressively tighter from left
to right.  We would like to conveniently wrap this domain-specific knowledge inside
our Bayesian framework later.  To this end, \cite{moran2023mira} 
used the multiple low- and high-resolution spectra across all cosmologies to
obtained precision estimates of these spectra across the $n$ 
wavenumber values ($p_1,\dots,p_n$) using a log-log regression model and a multiplier 
$c$ for the increase in precision from the low- to high- resolution output.
To streamline notation, we denote three unique diagonal matrices:
\[
\Lambda_p = 
   \begin{tikzpicture}[decoration={brace,amplitude=5pt},baseline=(current bounding box.west)]
     \matrix (magic) [matrix of math nodes,left delimiter=(,right delimiter=)] {
      10^5 \\
      & \ddots \\
      & & 10^5 \\
      & & & 0 \\
      & & & & 0 \\
      & & & & & \ddots \\
      & & & & & & 0\\
     };
     \draw[decorate] (magic-1-1.north) -- (magic-3-3.north east) node[above=5pt,midway,sloped] {$k < 0.04$};
   \end{tikzpicture}
\quad
\Lambda_l = 
   \begin{tikzpicture}[decoration={brace,amplitude=5pt},baseline=(current bounding box.west)]
     \matrix (magic) [matrix of math nodes,left delimiter=(,right delimiter=)] {
      0 \\
      & \ddots \\
      & & 0 \\
      & & & 10p_1 \\
      & & & & \ddots \\
      & & & & & 10p_n ? \\
      & & & & & & 0 \\
      & & & & & & & \ddots \\
      & & & & & & & & 0 \\
     };
     \draw[decorate] (magic-4-4.north) -- (magic-6-6.north east) node[above=5pt,midway,sloped] {$0.04\leq k < 0.25$};
   \end{tikzpicture}
\]
\[
\Lambda_h = 
   \begin{tikzpicture}[decoration={brace,amplitude=5pt},baseline=(current bounding box.west)]
     \matrix (magic) [matrix of math nodes,left delimiter=(,right delimiter=)] {
      0 \\
      & \ddots \\
      & & 0 \\
      & & & cp_1 \\
      & & & & \ddots \\
      & & & & & cp_n \\
     };
     \draw[decorate] (magic-4-4.north) -- (magic-6-6.north east) node[above=5pt,midway,sloped] {$0.25\leq k \leq 5$};
   \end{tikzpicture}
\]
\textcolor{red}{(@Steve - I am still quite confused here.  Are the above matrices helpful?
I'm confused by $p_i$.  Isn't there some overlap
between the low and high res runs?  The piecewise form you had doesn't have any overlap.  You
had the large $\Lambda$ as the concatenation of the matrices, but then you also had
it as a sum.  Perhaps we should talk this out sometime.  There has to be an easier 
way to convey how we calculate the weighted average.  Perhaps we could represent 
this better with indicator functions?  
Or as a function of $k_i$?  The equations below are just some ideas.)}
\[
\Lambda_p(k) = 10^5 \mathbbm{I}_{\{k < 0.04\}}
\]
where $\mathbbm{I}$ is the indicator function.
\[
\Lambda(k_i) = \begin{cases}
    10^5 \text{ for } k_i < 0.04 \text{  (perturbation theory)}\\
    16p_i \text{ for } 0.04 \leq k_i < 0.25 \text{  (low resolution)}\\
    cp_i \text{ for } 0.25 \leq k_i \leq 5 \text{  (high resolution)}
    \end{cases},
\]
where we use the sufficiently high precision of $10^5$ to impose the deterministic
nature of the perturbation theory output.  \textcolor{red}{(Then comment on how 
we can put this all into one big diagonal matrix that we multiply to our y values?)}

% $\Lambda_p = \text{diag}\left((10^5)_{k < 0.04}\right)$, 
% $\Lambda_{\ell} = \text{diag}\left(16\times(p_1,\dots,p_n)_{0.04 \leq k < 0.25}\right)$, 
% $\Lambda_h = \text{diag}\left(c\times(p_1,\dots,p_n)_{0.04 \leq k \leq 5}\right)$.

%\[
%\Lambda = 
%\begin{bmatrix}
%\Lambda_p & 0 & 0 \\
%0 & \Lambda_\ell & 0 \\
%0 & 0 & \Lambda_h
%\end{bmatrix}
%=
%\begin{bmatrix}
%\operatorname{diag}(10^5, \ldots, 10^5) &  &  \\
% & \operatorname{diag}(16\times(p_1, \ldots, p_n)) &  \\
% &  & \operatorname{diag}(c\times(p_1, \ldots, p_n))
%\end{bmatrix}
%\]

\textcolor{red}{(This part will need some revising, it is still unclear to me
how the weighted average is calculated.  Although I know in reality it is super simple.}
Above, each matrix element is either the corresponding precision value or zero, 
depending on whether the $k$ value is approximately unbiased for that data type. 
With this, we calculate a weighted average for each cosmology: 
$\bar y = \Lambda^{-1}(\Lambda_p y_p + \Lambda_h y_h + \Lambda_{\ell} \bar{y}_\ell)$, 
where $\Lambda = \Lambda_p + \Lambda_\ell + \Lambda_h$ is the precision matrix for 
the weighted average for a particular cosmology. On its own, the weighted average
has too much variation baked in from the limited number of model runs. There is also
no spatial dependence accounting for the correlated errors seen in Figure \ref{fig:plot_data}.
To account for these modeling challenges, we need a more complex model.

We consider a training-testing split with 111 cosmologies used for estimation and 6 
cosmologies held out for testing.  Our ultimate objective is to effectively predict
the underlying power matter spectra for the held-out cosmologies based on their 
8-dimensional input configuration (Section \ref{sec:pred}).  But first, we must 
obtain effective estimates of the underlying spectrum for each training 
cosmology (Section \ref{sec:hm_fit}).

>>>>

Notice, in the center panel of Figure \ref{fig:plot_data}, the precision of each
low/high resolution spectra increases as $k$ increases.  The curves get progressively 
tighter from left to right.  We would like to conveniently wrap this domain-specific 
knowledge inside our Bayesian framework later.  To this end, \cite{moran2023mira} 
used the multiple low- and high-resolution spectra across all cosmologies to
obtained precision estimates, $p_1,\dots, p_n$, across the $n$ 
wavenumber values using a log-log regression model.
% and a multiplier $c$ for the increase in precision from the 
% low- to high- resolution output.
We use these estimates to define three diagonal precision matrices, 
one for each data type (perturbation theory, low resolution, and high resolution).
Specifically, denote $\Lambda_p$, $\Lambda_\ell$, and $\Lambda_h$ as the
diagonal matrices with elements
\[
\Lambda_p^{(ii)} = \begin{cases}
    10^5 &\text{for } k_i < 0.04 \\
    0  &\text{otherwise}\\
    \end{cases}
\quad
\Lambda_\ell^{(ii)} = \begin{cases}
    16p_i &\text{for } 0.04 \leq k_i < 0.25 \\
    0  &\text{otherwise}\\
    \end{cases}
\quad
\Lambda_h^{(ii)} = \begin{cases}
    cp_i &\text{for } 0.04 \leq k_i < 5 \\
    0  &\text{otherwise}\\
    \end{cases}
\]
for $i=1,\dots, n$.  In $\Lambda_p$, we use the sufficiently high precision 
of $10^5$ to impose the deterministic nature of the perturbation theory output.  
\textcolor{red}{(@Steve - what is $c$
exactly?)}  We then calculate a weighted average for each cosmology: 
$\bar y = \Lambda^{-1}(\Lambda_p y_p + \Lambda_h y_h + \Lambda_{\ell} \bar{y}_\ell)$, 
where $\Lambda = \Lambda_p + \Lambda_\ell + \Lambda_h$ and 
$\bar{y}_\ell = \frac{1}{16}\sum_{r=1}^{16} y_{\ell_r}$. In Figure \ref{fig:plot_data},
the weighted average is shown in dashed orange.  As an estimate of the true
underlying spectra, the weighted average fluctuates too drastically due 
to the limited number of model runs.  It fails to account for the
spatial dependence inherent in the smooth but stochastic spectra.
To account for these modeling challenges, we need a more complex model.

We consider a training-testing split with 111 cosmologies used for estimation and 6 
cosmologies held out for testing.  Our ultimate objective is to effectively predict
the underlying power matter spectra for the held-out cosmologies based on their 
8-dimensional input configuration (Section \ref{sec:pred}).  But first, we must 
obtain effective estimates of the underlying spectrum for each training 
cosmology (Section \ref{sec:hm_fit}). \textcolor{red}{(@Steve - maybe this is a
good place to introduce the ``hold out'' and ``within sample'' terms?)}
